The need to reinterpret Hodgkins et al. (2018)

### 2022-11-23 README.Rmd
Rework manuscript 2, update the README.

> Henning Teickner, and Klaus-Holger Knorr (2022; preprint): _Improving Models to Predict Holocellulose and Klason Lignin Contents for Peat Soil Organic Matter with Mid Infrared Spectra_. Preprint. Soil and methods, June 1, 2022. DOI: [10.5194/soil-2022-27](https://doi.org/10.5194/soil-2022-27).

>>>>

> Henning Teickner, and Klaus-Holger Knorr (2022): _Improving Models to Predict Holocellulose and Klason Lignin Contents for Peat Soil Organic Matter with Mid-Infrared Spectra_. SOIL 8 (2): 699–715. DOI: [10.5194/soil-8-699-2022](https://doi.org/10.5194/soil-8-699-2022).

### 2022-11-23 analysis/paper/001-reply-main.Rmd
Rework manuscript 2, update the README.

---
title: 'Supplementary Information to: "Predicting Absolute Holocellulose and Klason Lignin Contents for Peat Remains Challenging"'
author: "Henning Teickner, Klaus-Holger Knorr"
date: ""
output: 
  bookdown::pdf_book:
    keep_tex: true
    citation_package: natbib
    toc: false
bibliography: references
biblio-style: unsrtnat
natbiboptions: super
header-includes:
   - \usepackage{float}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
---

@@

1. The models described in @Hodgkins.2018 to predict holocellulose and Klason lignin contents from mid infrared spectra, but recomputed in a Bayesian framework. This is the "original Bayesian model" in @Teickner.2022a. These models were used to reproduce the computations in @Hodgkins.2018. Here, these models are called "original" models.

2. A modified model to predict each, holocellulose and Klason lignin contents from mid infrared spectra. These are the models called "best binned spectra" models in @Teickner.2022a. A brief description of these models is given in section [Computation of modified models to predict organic matter holocellulose and Klason lignin contents from mid infrared spectra]. These models are used to reanalyze holocellulose and Klason lignin depth profiles for the two latitude classes defined in @Hodgkins.2018, as described in section [Depth profiles].

@@

Computation of the modified models using only the mid infrared spectra (MIRS) is described in detail in @Teickner.2022a. In brief, we used the same training data as @Hodgkins.2018 and a Bayesian linear regression model which we computed with Markov Chain Monte Carlo (MCMC) sampling using brms [@Burkner.2017; @Burkner.2018] and rstan [@StanDevelopmentTeam.2020] which are both interfaces to Stan [@StanDevelopmentTeam.2021]. In contrast to the original models of @Hodgkins.2018, we (1) used a beta distribution as likelihood function instead of assuming a Gaussian distribution, because holocellulose and Klason lignin contents cannot be negative or larger than 100 mass-%, we (2) did not use manually selected peaks extracted from the spectra, but all the $z$-transformed absorbance values after baseline correction, normalization, and binning (bin width = 20 cm$^{-1}$), and we (3) did not use simple linear regression, but regularized linear regression by using a regularized horseshoe prior [@Piironen.2017a; @Piironen.2017b] for coefficients for spectral variables. For the intercept and for the scale parameter of the beta distribution, we used weakly informative default priors of rstanarm [@Goodrich.2020] and brms [@Burkner.2017; @Burkner.2018].

@@

1. Since representative training data are missing, both the original and modified models cannot be fully validated and prediction uncertainties are probably larger than computed with these models [@Teickner.2022a]. Since representative test data are missing, these additional uncertainties cannot be estimated. We assume that these uncertainties are larger than those induced by parameter uncertainty from models for the depth profiles.

@@

The modified models using only the MIRS were also used to compute median values and confidence intervals mentioned in the main text in section "Klason lignin content and residual enrichment in (sub)tropical peatlands probably are larger". In these computations, we also considered the predictive uncertainty of the original models (the Bayesian formulation described in @Teickner.2022a).

@@

- $\text{Y}_1$ are the measured holocellulose contents of the training data.
- $\mu_{1}$ are the respective average values modeled with MIRS.
- $\text{X}_1$ is a matrix containing the MIRS variables (variables in columns and samples in rows) for the training data. 
- $i$ is an index for each sample of the training data.
- $\phi$ is the precision parameter for the Beta distribution with which values of $\text{Y}_1$ are modeled.
- $\alpha$ and $\beta$ are intercept and regression coefficients for the MIRS prediction model. The $\beta$s are modeled via a regularized horseshoe prior parameterized following @Piironen.2017b as described in @Teickner.2022a.

@@

To create figure 1<!--\@ref(fig:reply-res-p-sites-latitude)--> in the main text, we used the models for the latitudinal gradient and samples from the posterior distribution to compute the predicted average and 95% confidence interval for a range of latitude values and we added predictions for the average contents for each of the sampled cores. Next, we recomputed the models from @Hodgkins.2018 by (1) predicting average peat sample contents with the original models (the Bayesian version described in @Teickner.2022a), (2) averaging the near surface peat sample average contents for each core, and (3) fitting an ordinary least squares regression line to these average core values. For this linear model, we plotted the confidence interval across latitude and we added the computed average core near-surface contents. Thus, figure 1<!---\@ref(fig:reply-res-p-sites-latitude)---> in the main text compares the graphed values and model prediction from figure 3 in @Hodgkins.2018 with predictions from our modified analysis considering prediction uncertainty from the MIRS-based model part, and the averaging per core.  
Additional sources for differences in predictions and prediction uncertainties in comparison to the analysis in @Hodgkins.2018 are the use of a beta distribution which constraints uncertainty intervals for extreme (near 0 and 100 mass-%) values, and the use of all spectral variables which resulted in reduced prediction uncertainties [@Teickner.2022a].

>>>>

---
title: 'Supplementary Information to: "The need to reinterpret Hodgkins et al. (2018)"'
author: "Henning Teickner, Klaus-Holger Knorr"
date: ""
output: 
  bookdown::pdf_book:
    keep_tex: true
    citation_package: natbib
    toc: false
bibliography: references
biblio-style: unsrtnat
natbiboptions: super
header-includes:
   - \usepackage{float}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
fontsize: 12pt
---

@@

1. The models described in @Hodgkins.2018 to predict holocellulose and Klason lignin contents from mid infrared spectra, but recomputed in a Bayesian framework. This is the "original Bayesian model" in @Teickner.2022f. These models were used to reproduce the computations in @Hodgkins.2018. Here, these models are called "original" models.

2. A modified model to predict each, holocellulose and Klason lignin contents from mid infrared spectra. These are the models called "best binned spectra" models in @Teickner.2022f. A brief description of these models is given in section [Computation of modified models to predict organic matter holocellulose and Klason lignin contents from mid infrared spectra]. These models are used to reanalyze holocellulose and Klason lignin depth profiles for the two latitude classes defined in @Hodgkins.2018, as described in section [Depth profiles].

@@

Computation of the modified models using only the mid infrared spectra (MIRS) is described in detail in @Teickner.2022f. In brief, we used the same training data as @Hodgkins.2018 and a Bayesian linear regression model which we computed with Markov Chain Monte Carlo (MCMC) sampling using brms [@Burkner.2017; @Burkner.2018] and rstan [@StanDevelopmentTeam.2020] which are both interfaces to Stan [@StanDevelopmentTeam.2021]. In contrast to the original models of @Hodgkins.2018, we (1) used a beta distribution as likelihood function instead of assuming a Gaussian distribution, because holocellulose and Klason lignin contents cannot be negative or larger than 100 mass-%, we (2) did not use manually selected peaks extracted from the spectra, but all the $z$-transformed absorbance values after baseline correction, normalization, and binning (bin width = 20 cm$^{-1}$), and we (3) did not use simple linear regression, but regularized linear regression by using a regularized horseshoe prior [@Piironen.2017a; @Piironen.2017b] for coefficients for spectral variables. For the intercept and for the scale parameter of the beta distribution, we used weakly informative default priors of rstanarm [@Goodrich.2020] and brms [@Burkner.2017; @Burkner.2018].

@@

1. Since representative training data are missing, both the original and modified models cannot be fully validated and prediction uncertainties are probably larger than computed with these models [@Teickner.2022f]. Since representative test data are missing, these additional uncertainties cannot be estimated. We assume that these uncertainties are larger than those induced by parameter uncertainty from models for the depth profiles.

@@

The modified models using only the MIRS were also used to compute median values and confidence intervals mentioned in the main text in section "Klason lignin content and residual enrichment in (sub)tropical peatlands probably are larger". In these computations, we also considered the predictive uncertainty of the original models (the Bayesian formulation described in @Teickner.2022f).

@@

- $\text{Y}_1$ are the measured holocellulose contents of the training data.
- $\mu_{1}$ are the respective average values modeled with MIRS.
- $\text{X}_1$ is a matrix containing the MIRS variables (variables in columns and samples in rows) for the training data. 
- $i$ is an index for each sample of the training data.
- $\phi$ is the precision parameter for the Beta distribution with which values of $\text{Y}_1$ are modeled.
- $\alpha$ and $\beta$ are intercept and regression coefficients for the MIRS prediction model. The $\beta$s are modeled via a regularized horseshoe prior parameterized following @Piironen.2017b as described in @Teickner.2022f.

@@

To create figure 1<!--\@ref(fig:reply-res-p-sites-latitude)--> in the main text, we used the models for the latitudinal gradient and samples from the posterior distribution to compute the predicted average and 95% confidence interval for a range of latitude values and we added predictions for the average contents for each of the sampled cores. Next, we recomputed the models from @Hodgkins.2018 by (1) predicting average peat sample contents with the original models (the Bayesian version described in @Teickner.2022f), (2) averaging the near surface peat sample average contents for each core, and (3) fitting an ordinary least squares regression line to these average core values. For this linear model, we plotted the confidence interval across latitude and we added the computed average core near-surface contents. Thus, figure 1<!---\@ref(fig:reply-res-p-sites-latitude)---> in the main text compares the graphed values and model prediction from figure 3 in @Hodgkins.2018 with predictions from our modified analysis considering prediction uncertainty from the MIRS-based model part, and the averaging per core.  
Additional sources for differences in predictions and prediction uncertainties in comparison to the analysis in @Hodgkins.2018 are the use of a beta distribution which constraints uncertainty intervals for extreme (near 0 and 100 mass-%) values, and the use of all spectral variables which resulted in reduced prediction uncertainties [@Teickner.2022f].

### 2022-11-23 analysis/paper/002-reply-main.Rmd
Rework manuscript 2, update the README.

---
title: 'Predicting Absolute Holocellulose and Klason Lignin Contents for Peat Remains Challenging'
author: "Henning Teickner$^{1, *}$, Klaus-Holger Knorr$^1$"
date: ""
output: 
  bookdown::pdf_book:
    keep_tex: true
    citation_package: natbib
    toc: false
bibliography: references
biblio-style: unsrtnat
natbiboptions: super
header-includes:
   - \usepackage{float}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
---

@@

The resilience of the large global peatland carbon stocks against climate change may depend on how peat chemistry controls decomposition rates [@Wang.2015; @Hodgkins.2018]. In their study, @Hodgkins.2018 argue that there is a latitudinal gradient in peat recalcitrance which may explain the stability of tropical peat in spite of warmer temperatures and which points toward a susceptibility of labile surface peat at high latitudes to degradation, but not deeper peat which has a chemistry more similar to tropical peat. This gradient can be described by differences in holocellulose contents --- representing relatively easily degradable carbohydrates --- and Klason lignin contents --- representing aromatics --- which @Hodgkins.2018 predicted from mid infrared spectra [@Hodgkins.2018].  
We acknowledge this conceptual model as an important step in understanding the stability of peat deposits and their susceptibility to degradation. However, we challenge the accuracy of the models used to predict peat holocellulose and Klason lignin contents. This inaccuracy has two major implications:

1. Peat Klason lignin contents and relative accumulation during decomposition probably are underestimated by @Hodgkins.2018. This may have important consequences for decomposition of high-latitude peat deposits due to climate change.

2. The inaccuracies imply that the predicted values are only very rough estimates for holocellulose and Klason lignin contents and at least partly biased. An interpretation of the values given in @Hodgkins.2018 in terms of quantitative concepts for organic matter chemistry (i.e. holocellulose and Klason lignin) therefore is not possible.  
Instead, we assume that if the study is reinterpreted in terms of qualitative concepts for carbohydrates and aromatics as used in the qualitative interpretation of mid infrared spectra for peat, there still is evidence for general latitudinal patterns in peat recalcitrance. However, interpretation and generalization of these findings are less straightforward.

Moreover, we suggest improvements for uncertainty propagation for the models. Remaining open questions to this approach of analyzing global peat chemistry patterns should be addressed in future studies.

## Patterns in peat chemistry should not be interpreted in terms of holocellulose and Klason lignin contents {.unnumbered}

In a recent evaluation of the models to predict peat holocellulose and Klason lignin contents presented in @Hodgkins.2018, we found that the training data used for the models (leaves, paper products) are not representative for peat [@Teickner.2022a].  
The peat samples are more decomposed than the training data samples. Decomposition typically causes distinct features in mid infrared spectra which have complex effects on the applicability of spectral prediction models [@Teickner.2022a]. Since these features are not present in the training data, predictions of the original models are extrapolations. Our re-evaluation showed that this causes large uncertainties in model predictions for both holocellulose and Klason lignin, and a comparatively large bias towards smaller Klason lignin contents, especially for more decomposed peat with Klason lignin contents >30 mass-% [@Teickner.2022a]. Because of these uncertainties and biases, the analyses of @Hodgkins.2018 cannot be interpreted in terms of absolute holocellulose and Klason lignin contents.  
Instead, we suggest to interpret differences in peat chemistry presented by @Hodgkins.2018 not in terms of the quantitative concepts (i.e. holocellulose and Klason lignin), but rather in terms of more qualitative concepts typically used in studies interpreting mid infrared spectra of peat: as "carbohydrates" and "aromatics" (e.g. @Artz.2008; @Broder.2012; @Tfaily.2014; @Biester.2014). We assume that under such an interpretation, the conclusions of @Hodgkins.2018 probably are valid.  
The proposed reinterpretation is possible if values given in @Hodgkins.2018 are only interpreted in terms of the normalized peak heights used in the original prediction models, and not as absolute holocellulose and Klason lignin contents. Each model is a linear model with one independent variable (the normalized height of one peak); therefore a direct reinterpretation of the predicted absolute contents in terms of normalized peak heights is possible if the figure axes are relabeled as "Relative content of carbohydrates/aromatics" instead of "Estimated % carbohydrates/aromatics" (implying an absolute content of holocellulose and Klason lignin, respectively) [@Hodgkins.2018].   
An important open question to this approach of interpreting large-scale gradients in peat chemistry based on mid infrared spectra should be addressed in future studies: Are differences in specific mid infrared spectra peaks reliable indicators for differences in a specific concept for peat chemistry? For example, are differences in the peak around 1040 cm$^{-1}$ indicative for differences in carbohydrate contents (also across larger scales)?  
A similar problem arose in the discussion on the use of the C/N ratio as a good indicator for peat mass loss. This may only hold true if there have been no or negligible shifts in the vegetation composition which causes differences in the C/N ratio already in initial plant litter (e.g. @Biester.2014). For mid infrared spectra, such knowledge still has to be acquired in future studies, especially if large environmental gradients are considered as in @Hodgkins.2018.

## Klason lignin content and residual enrichment in (sub)tropical peatlands probably are larger {.unnumbered}

We suggest that @Hodgkins.2018 have underestimated peat Klason lignin contents, especially in (sub)tropical peat. As described above, the original model for Klason lignin underestimates peat Klason lignin contents. A modified model has less bias [@Teickner.2022a]. In spite of several remaining limitations (see below and the supplementary information), we argue that this modified model can serve as rough indicator for the underestimation of peat Klason lignin contents by the original model. We therefore present values to provide tentative estimates for the underestimation of Klason lignin contents, but we do not claim that these estimates are optimal.  
With this modified model, near-surface ($\le50$ cm) peat Klason lignin contents are on average ~10 to 15 mass-% larger across the latitudinal gradient (figure \@ref(fig:reply-res-p-sites-latitude)). Specifically, average (sub)tropical near-surface peat ($<45 ^\circ$N, $\le50$ cm) Klason lignin contents are `r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% (median, lower and upper 95% prediction interval limit) larger than with the original model. There is also tentative indication for on average larger differences between deep (sub)tropical ($<45 ^\circ$N) and high-latitude ($\ge45 ^\circ$N) peat (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).    
We also suggest that the residual enrichment of Klason lignin during decomposition probably has been underestimated. A rough estimate for the residual enrichment of Klason lignin during decomposition is the difference in Klason lignin content between near-surface peat ($\le50$ cm) and deeper peat. With the modified model, this difference is on average for (sub)tropical peatlands `r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% larger (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). For high latitude peatlands the difference is smaller on average (`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-%).  
Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this in light of climate change is that high latitude peat deposits may be more prone to shifts in decomposition rates than suggested by the original results.

## Larger uncertainties due to a lack of uncertainty propagation and unrepresentative data {.unnumbered}

Besides a reinterpretation in terms of qualitative concepts for organic matter recalcitrance, we also suggest that a more comprehensive quantification and propagation of uncertainties is important to further scrutinize the results of @Hodgkins.2018 because these uncertainties are not negligibly small.  
In the original analysis, the following uncertainties are omitted: (1) As described above, the training dataset is not representative for peat and therefore predictions are biased. (2) The prediction uncertainty of the models used to predict holocellulose and Klason lignin contents. (3) In computing the regression model to analyze the latitudinal gradient based on average core near-surface contents, the uncertainty for these average values was not propagated.  
The first two sources of uncertainty could be omitted if only normalized peak heights (as qualitative indicators) are used instead of predicted holocellulose and Klason lignin contents, as described above. But the third source of uncertainty should also be considered when reinterpreting the results in terms of area normalized peak heights.  
To support our concerns, we reanalyzed the latitudinal gradient for holocellulose and Klason lignin, respectively, using the modified models described in @Teickner.2022a and fully considering the last two sources of uncertainty using a Bayesian approach (described in more detail in the supplementary information). With the modified models, the estimated slopes for the latitudinal gradient are more uncertain (95% confidence intervals are [`r s1_res_holocellulose %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] and [`r s1_res_klason_lignin %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] for holocellulose and Klason lignin, respectively; figure \@ref(fig:reply-res-p-sites-latitude) and supporting figure 3<!---\@ref(fig:reply-res-p-latitude-poc-slope)--->). A similar reanalysis of the depth profiles between the two latitude categories shows that differences in depth patterns are also more uncertain (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).  
Our analysis therefore indicates larger uncertainties both in the depth profiles and the latitudinal gradient. Considering these uncertainties is important when peat properties predicted with a model are analyzed.
Again, the values presented here are only tentative because the modified models probably are also biased.

## Future directions {.unnumbered}

Notwithstanding our critical remarks, the idea of @Hodgkins.2018 to relate mid infrared spectra to precisely quantifiable and defined organic matter fractions is important. Such fractions are relatively well understood and therefore allow us to better deduce what effects climate shifts can have on peat decomposition and recalcitrance: It is much more informative if we are able to precisely estimate a change in Klason lignin contents of e.g. 5 mass-% if average annual temperatures increase by 1$^{\circ}$C than if we are only able to make qualitative statements that the relative amount of aromatics (without describing in detail what this fraction means) increases, as is currently the case.  
When interpreting mid infrared spectra, distinguishing between precisely quantifiable concepts for organic matter chemistry, such as holocellulose and Klason lignin contents, and qualitative concepts which currently are used in interpreting peat MIRS, such as "carbohydrates and aromatics" (without specifying how this fraction could actually be measured), is crucial: Models relying on quantitative concepts are more testable and allow to actually quantify fractional changes in organic matter chemistry, whereas models built on qualitative concepts can only be refuted if there are blatant discrepancies relative to expectations (i.e. a decrease in "carbohydrates" with latitude). Ultimately, we need more precise models for variations in peat chemistry which can be developed based on qualitative concepts.  
We suggest the following steps to gain deeper insights into latitudinal gradient in peat chemistry: As outlined above, we suggest to interpret the results of @Hodgkins.2018 in terms of qualitative concepts for peat chemistry, i.e. "carbohydrates" and "aromatics". These suggestions also apply to a follow-up study using the same models [@Verbeke.2022]. Further studies are needed to check if specific peaks are good indicators for these qualitative concepts for peat chemistry, how the relation may be confounded e.g. by vegetation composition or other organic matter fractions, as could be the case especially for peaks related to aromatics [@Teickner.2022a]. This could be done by computing models to reliably predict holocellulose and Klason lignin contents of peat from mid infrared spectra and analyzing which spectral variables are relevant for this purpose and if there are confounding variables. Such models will also help us to get more precise insights into the degree to which high latitude peat accumulation may be decreased by decomposition due to rising temperatures, because they allow replacing more costly measurements. Finally, future studies should carefully consider uncertainties and fully propagate these into model estimates if it is unknown whether such uncertainties are negligible. Predicting absolute peat holocellulose and Klason lignin contents from infrared spectra remains challenging.

@@

We like to thank Jeffrey Chanton and Suzanne Hodgkins for a constructive dicussion which greatly helped to improve a first version of this manuscript. This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) grant no. KN 929/23-1 to Klaus-Holger Knorr and grant no. PE 1632/18-1 to Edzer Pebesma.

>>>>

---
title: 'The need to reinterpret Hodgkins et al. (2018)'
author: "Henning Teickner$^{1, *}$, Klaus-Holger Knorr$^1$"
date: ""
output: 
  bookdown::pdf_book:
    keep_tex: true
    citation_package: natbib
    toc: false
bibliography: references
biblio-style: unsrtnat
natbiboptions: super
header-includes:
   - \usepackage{float}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
fontsize: 12pt
---

@@

@Hodgkins.2018 used spectral prediction models to estimate two major peat organic matter fractions --- holocellulose and Klason lignin --- in peat cores collected along a latitudinal gradient. Based on this, @Hodgkins.2018 argue that (sub)tropical peat deposits accumulate despite high temperatures due to higher Klason lignin over holocellulose contents, and that climate warming is a risk to northern surface peat deposits which would decompose faster due to higher holocellulose contents, but not to deeper peat which is more similar to tropical peat. In a recent study, we have shown that the spectral prediction models behind this conclusion are not valid for peat samples [@Teickner.2022f]. In addition to that, we show here that @Hodgkins.2018 did not consider important uncertainty sources in their analysis. This raises the question whether the results of @Hodgkins.2018, and of a follow-up study with more data [@Verbeke.2022], remain valid or how they have to be reinterpreted.

We show that the results of @Hodgkins.2018 are to some extent still valid for holocellulose, but only when reinterpreted in terms of differences in peak heights, which is a major obstacle to a quantitative understanding of the control of peat chemistry on decomposition. For Klason lignin, remaining uncertainties are high even after reinterpretation in terms of peak heights. Even more, there is evidence that near-surface peat Klason lignin contents and residual enrichment of Klason lignin during decomposition have been underestimated. It thus remains uncertain whether and to what extent the reported latitudinal gradient in peat chemistry will control future peatland carbon losses.

## Why are the results highly uncertain? {.unnumbered}

The original models to estimate peat holocellulose and Klason lignin contents are not valid and, highly uncertain, and thus likely biased --- especially for decomposed peat --- and therefore should not be used to estimate holocellulose and Klason lignin contents [@Teickner.2022f]. 
Besides, there are that two important sources of uncertainty have been neglected in @Hodgkins.2018: first, the prediction uncertainty of the spectral prediction models, and second, the uncertainty introduced when computing near-surface average holocellulose and Klason lignin contents from estimates for individual peat layers.  
We illustrate the impact of all these factors by recomputing the latitudinal gradient and depth profiles with the modified, probably more accurate, but still biased, models from @Teickner.2022f, while considering these uncertainty sources (supporting information S1).  
This analysis shows that both the estimated slopes for the latitudinal gradient (95% confidence intervals are [`r s1_res_holocellulose %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] and [`r s1_res_klason_lignin %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] for holocellulose and Klason lignin, respectively; figure \@ref(fig:reply-res-p-sites-latitude) and supporting figure 3<!---\@ref(fig:reply-res-p-latitude-poc-slope)--->) and differences in the depth profiles are highly uncertain due to these uncertainty sources (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).

## Klason lignin contents and residual enrichment with decomposition probably are larger {.unnumbered}

In a previous publication, we have shown that the spectral prediction model for Klason lignin underestimates Klason lignin contents especially when these are $>30$ mass-% [@Teickner.2022f]. Our modified model is much less biased [@Teickner.2022f] and therefore enables a rough analysis of the order of magnitude of the bias. This allowed us to analyze what consequences the bias can have for the estimated near-surface ($\le50$ cm) peat Klason lignin contents and residual enrichment of Klason lignin during decomposition (supporting information S1).   
With our modified model, near-surface peat Klason lignin contents are on average ~10 to 15 mass-% larger across the latitudinal gradient (figure \@ref(fig:reply-res-p-sites-latitude)). Specifically, average (sub)tropical ($<45 ^\circ$N) near-surface peat Klason lignin contents are `r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% (median, lower and upper 95% prediction interval limit) larger than with the original model. There is also tentative indication for on average larger differences between deep (sub)tropical ($<45 ^\circ$N) and high-latitude ($\ge45 ^\circ$N) peat (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).    
Residual enrichment of Klason lignin during decomposition probably has been underestimated. A rough estimate for the residual enrichment of Klason lignin during decomposition is the difference in Klason lignin content between near-surface peat and deeper peat. With the modified model, this difference is on average for (sub)tropical peatlands `r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% larger (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). For high latitude peatlands the difference is smaller (`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-%).  
Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this in light of climate change is that high latitude peat deposits may be even more prone to shifts in decomposition rates than suggested by the original results.

## How to reinterpret @Hodgkins.2018 {.unnumbered}

The mid-infrared spectra (MIRS), which have been used to compute the spectral prediction models, have so far mostly been interpreted in terms qualitative concepts for organic matter fractions (e.g. @Cocozza.2003): the `carb` peak [@Hodgkins.2018] --- the only predictor for holocellulose content in @Hodgkins.2018 --- is an indicator for "carbohydrates" in mineral-poor peat [@Hodgkins.2018]; the `arom15arom16` peak [@Hodgkins.2018] --- the only predictor for Klason lignin content in @Hodgkins.2018 --- is an indicator for "aromatics" in peat with low protein and carbonate contents [@Stuart.2004; @Tatzber.2007; @Teickner.2022f] (note that we write "carbohydrates" and "aromatics" because, when interpreted only in terms of peak heights, it is remains uncertain what these organic matter (OM) fractions represent chemically or how one could extract them).

Such a qualitative interpretation of MIRS enables us to reinterpret @Hodgkins.2018, circumventing some of the uncertainties we have uncovered: All holocellulose and Klason lignin values given in @Hodgkins.2018 are directly proportional to the heights of the `carb` and `arom15arom16` peak, respectively. This removes uncertainties due to invalid spectral prediction models from the analyses. Then, we can still say that @Hodgkins.2018 show systematic latitudinal and depth-related differences in peat "carbohydrate" and "aromatics" contents.

The trade-offs are first, that we have to make the assumption that there are no latitudinal or depth-related differences in mineral, carbonate, and protein contents, and second, that we have to make the assumption that differences in heights of the two peaks are not dominated by differences in other peaks in the spectra. The reason is that differences in peak heights can be attributed to "carbohydrates" or "aromatics" only when contents of minerals or proteins and carbonates, respectively, are the same, and when changes in the area normalized height of these peaks are not caused by changes in other peaks, which can be an effect of spectral normalization [@Stuart.2004; @Tatzber.2007; @Hodgkins.2018; @Teickner.2022f].

We think that only some of these assumptions can are supported at the moment. The `carb` peak is comparatively large and has low interferences with peaks caused by other molecular structures. It should therefore be a good indicator for "carbohydrates" because mineral contents in (near-surface) peat are typically low (as validated by @Hodgkins.2018). The `arom15arom16` peak has large contributions by proteins [@Stuart.2004; @Teickner.2022f], based on peat N contents, it is not unlikely that peat protein contents differ with depth and latitude [@Sowden.1978; @Loisel.2014], and it may be more affected by other peaks since it is smaller than the `carb` peak. For this reason, we suggest that more evaluation is necessary before accepting, based on @Hodgkins.2018, that there are systematic differences in peat "aromatic" contents.

The reinterpretation in terms of "carbohydrate" and "aromatics" instead of holocellulose and Klason lignin has the clear disadvantage that we sacrifice the quantitative precision @Hodgkins.2018 aimed to attribute to the results: it is much more informative to precisely estimate a change in Klason lignin content of e.g. 5 mass-% when the average annual temperature increase by 1$^{\circ}$C than to only make qualitative statements that the relative amount of "aromatics" (without describing in detail what this OM fraction means) increases.

We think @Hodgkins.2018 have advanced a very important idea to make the interpretation of MIRS more quantitative. Only with such a quantitative approach, which links different types of measurements and the knowledge we have about them, can we develop a quantitatively precise understanding. However, we have shown that there is some more way to go. We hope that our critical analysis will help to improve this approach.

@@

We like to thank Jeffrey Chanton and Suzanne Hodgkins for a constructive discussion which greatly helped to improve a first version of this manuscript. This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) grant no. KN 929/23-1 to Klaus-Holger Knorr and grant no. PE 1632/18-1 to Edzer Pebesma.
